---
title: "Bud segmentation of mRNAs"
author: "Edward Wallace"
date: "2026-07-13"
output: 
  html_document:
      toc: true
---

# Summary

Analysis of CLB2 + ASH1, SUN4, SRL1, mRNAs by smFISH including bud segmentation.
Population-level cell analysis in R started by EW on 8 July 2026, continued through 13 July 2026.

Goal is to produce figures summarising the dataset:

- abundance and bud localization of mRNAs in wild-type
- comparative bud localization in different strains
- whatever cell statistics we need to support these arguments.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(ggbeeswarm)
library(GGally)
library(cowplot)
theme_set(theme_cowplot(font_size = 10) + 
            theme(strip.background = element_blank())
)
```


# Load bud segmentation data

Define pixel dimensions in micrometres (um), 0.065 micron per pixel for the 100x objective.

Also set range of sizes to plot for total cell area, bud area, nuclear area, axis limits, in micrometres.

```{r pixel_dimensions}
pixel_width <- 0.065
pixel_area  <- pixel_width^2

cell_area_limits <- c(6,30)
bud_area_limits  <- c(0,15)
nuc_area_limits  <- c(0,3.5)

cell_length_limits <- c(2, 20)
```

```{r strain_names}
# make short table here of strain ids and names in desired order
strain_table <-
  tribble(~strain, ~condition, ~strain_nice,
    "YET915", "WT",    "WT",
    "YET916", "ΔSSD1", "ssd1Δ",
    "YET919", "ΔMPT5", "mpt5∆",
    "YET563", "ΔSHE2", "she2∆",
    "YET572", "ΔSHE3", "she3∆",
    "YET918", "ΔSUN4", "sun4∆",
    "YET921", "ΔSEC72","sec72∆"
    ) %>%
  mutate(strain_nice = as_factor(strain_nice))

strain_table
```



```{r mrna_names}
mrna_names <- c("CLB2 + ASH1", "SRL1", "SUN4")

mrna_names

scale_colour_mrna <- function(...) {
  scale_colour_manual(
    values = c("CLB2 + ASH1" = "magenta3", 
               "SRL1" = "goldenrod3", 
               "SUN4" = "forestgreen"),
    ...)
}
```

```{r load_data}
mrna_alldata <- here::here("data", "Mara_budSegmentationData20260311.csv") %>%
  read_csv() %>%
  select(!starts_with("Unnamed") ) %>%
  select(!"condition + strain") %>%
  rename(row_id = "...1") %>%
  mutate(replicate = 
           factor(replicate,
                  levels = c("BR1", "BR2", "BR3", "BR4", "BR5")),
         mRNA = factor(mRNA,
                       levels = c("ASH1CLB2", "SRL1", "SUN4"), 
                       labels = c("CLB2 + ASH1", "SRL1", "SUN4"), ),
         area_umsq     = area * pixel_area,
         area_bud_umsq = area_bud * pixel_area,
         axis_minor_length_um = axis_minor_length * pixel_width,
         axis_major_length_um = axis_major_length * pixel_width,
         ) %>%
  rename_with(str_replace_all, pattern = "-", replacement = "_") %>%
  left_join(strain_table, by = c("strain", "condition"))

mrna_alldata
```


```{r count_by_replicate}
mrna_alldata %>% 
  filter(mRNA == "SRL1") %>%
  group_by(strain_nice, replicate) %>% 
  count() %>%
  print(n = 30)
```

# Overview plots across strains

## mRNAs per cell

```{r plot_mrna_per_cell, fig.height = 4, fig.width = 4.5}

ggplot(data = mrna_alldata) +
  geom_boxplot(aes(x = strain_nice, colour = mRNA, y = total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```


## mRNAs per mother

```{r plot_mrna_per_mother_allstrains, fig.height = 4, fig.width = 4.5}

ggplot(data = mrna_alldata, aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_mother),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```

## mRNAs per bud, filtering only for budded cells

```{r plot_mrna_per_bud_allstrains, fig.height = 4, fig.width = 4.5}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0), 
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```


## Count vs total cell area for WT 

```{r plot_countvsarea_WT, fig.height = 4, fig.width = 3.5}
plot_count_vs_area_WT <- 
  ggplot(data = mrna_alldata %>%
           filter(area_umsq > 4,
                  strain_nice %in% c("WT")),
         aes(x = area_umsq, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "gam", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  coord_cartesian(expand = FALSE) + 
  scale_colour_mrna() +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  labs(x = "total cell area (um^2)",
       y = "mRNA count per cell") + 
  guides(colour = guide_legend(order = 1),
         linetype = "none")

plot_count_vs_area_WT
```

```{r plot_countvsarea_vsstrain, fig.width = 7}
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆"),
                area_bud > 0),
       aes(x = area_umsq, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(colour = mRNA),
              method = "gam",  se = FALSE) + 
  facet_grid(mRNA ~ strain_nice, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  coord_cartesian(xlim = cell_area_limits, expand = FALSE) +
  scale_colour_mrna() + 
  labs(x = "total cell area (um^2)",
       y = "mRNA count per cell")

```

## Count vs bud area for WT

```{r plot_countvsbudarea_WT, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0,
                strain_nice %in% c("WT")),
       aes(x = area_bud_umsq, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "gam", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  scale_colour_mrna()  +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  labs(x = "bud area (um^2)",
       y = "mRNA count per cell")

```


## Proportion in bud, filtering only for budded cells 

```{r plot_propinbud_allstrains, fig.height = 4, fig.width = 5}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous(limits = c(0,1)) +
  scale_colour_mrna()

```

## Proportion in bud, across biological replicates, for most relevant strains 

```{r plot_propinbud_replicates, fig.height = 4, fig.width = 4}
plot_propinbud_replicates <- 
  ggplot(data = mrna_alldata %>%
           filter(area_bud > 0,
                  strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆")),
         aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs, 
                   group = interaction(strain_nice, replicate)),
               outlier.size = 0.2,
               position = position_dodge(preserve = "single")) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNAs in bud",
                     limits = c(0,1)) +
  scale_colour_mrna() + 
  labs(x = "Strain")

plot_propinbud_replicates

```


## Prop in bud vs bud size for WT

```{r plot_propinbud_WT, fig.height = 4, fig.width = 3.5}
plot_prop_vs_budarea_WT <-
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT"),
                area_bud > 0),
       aes(x = area_bud_umsq, y = total_RNAs_bud / total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "gam", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNA in bud", 
                     limits = c(0,1)) +
  labs(x = "Bud area (um^2)") + 
  coord_cartesian(xlim = bud_area_limits, expand = FALSE) +
  scale_colour_mrna() + 
  guides(colour = guide_legend(order = 1),
         linetype = "none")

plot_prop_vs_budarea_WT
```



## Prop in bud vs bud size for relevant strains

```{r plot_propinbud_vsstrain, fig.height = 4, fig.width = 6}
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆"),
                area_bud > 0),
       aes(x = area_bud_umsq, y = total_RNAs_bud / total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(colour = mRNA),
              method = "gam",  se = FALSE) + 
  facet_grid(mRNA ~ strain_nice) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNA in bud",
                     limits = c(0,1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c("0", "", "0.5", "", "1")) +
  labs(x = "Bud area (um^2)") + 
  coord_cartesian(xlim = bud_area_limits, expand = FALSE) +
  scale_colour_mrna()

```


## Distribution of cell area by strain

```{r plot_cellareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata,
       aes(x = area_umsq)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = cell_area_limits, expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of bud area by strain

```{r plot_budareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = area_bud)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = bud_area_limits, expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of mother nuclear area by strain

```{r plot_mothernucareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata ,
       aes(x = mother_nuclei_area)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = nuc_area_limits, expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of bud nuclear area by strain

Looks terrible because most buds do not have nuclei.

```{r plot_budnucareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = nuclei_bud_area)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  #coord_cartesian(xlim = c(0,750), expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

# Assign cell cycle phase based on cell statistics

The idea is:

- G1 cells have no bud and nucleus in mother
- S cells have small buds (under about 20 um^2)
- G2 cells have larger buds but only one nucleus in mother
- M cells have 2 nuclei in both bud and mother
- some cells might be unassigned to "other"

```{r assign_cell_phase}
mrna_alldata_wphase <-
  mrna_alldata %>%
  # filter for cells of sensible size - 2 um to 20 um
  filter(axis_major_length_um > cell_length_limits[1],
         axis_major_length_um < cell_length_limits[2]) %>%
  mutate(
    cell_cycle_phase = 
      case_when(
        area_bud_umsq == 0 ~ 
          "G1",
        area_bud_umsq > 0    & 
          area_bud_umsq < 2  & 
          nuclei_bud_count == 0 ~ 
          "S",
        area_bud_umsq >= 2 & 
          area_bud_umsq < bud_area_limits[2] & 
          nuclei_bud_count == 0 ~ 
          "G2",
        area_bud_umsq > 2  & 
          mother_nuclei_count == 1 & 
          nuclei_bud_count == 1 ~ 
          "M",
        .default = "NA"
      ) %>%
      factor(levels = c("G1", "S", "G2", "M"))
  )
```

We should check what the unassigned cells are doing - cells that are unreasonably small or have unreasonably large buds - but it is not a large proportion.


### Check cell-level statistics for cell cycle phase.

```{r check_stats_cell_cycle_phase, fig.height = 7, fig.width = 7}
mrna_alldata_wphase %>%
  filter(mRNA == "SUN4") %>%
  mutate(axis_major_length_bud_um = 
           axis_major_length_bud * pixel_width ) %>%
  select(area_umsq,
         axis_major_length_um, 
         area_bud_umsq,
         axis_major_length_bud_um, 
         cell_cycle_phase) %>%
  ggpairs(aes(colour = cell_cycle_phase), 
          lower = list(continuous = wrap("points", size = 0.2), 
                       combo = wrap("box_no_facet", outlier.size = 0.2),
                       discrete = "count") )
```

### Count mRNA by cell cycle phase

```{r countmrna_by_cellcycle_phase, fig.height=4, fig.width=3.5}
ggplot(data = mrna_alldata_wphase %>%
         filter( strain_nice == "WT")) +
  geom_boxplot(aes(x = cell_cycle_phase, colour = mRNA, y = total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()
```

### localise mRNA by cell cycle phase

```{r propmrna_by_cellcycle_phase, fig.height=4, fig.width=3.5}
ggplot(data = mrna_alldata_wphase %>%
         filter( strain_nice == "WT")) +
  geom_boxplot(aes(x = cell_cycle_phase, colour = mRNA, 
                   y = total_RNAs_bud / total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna() +
  scale_y_continuous("Proportion of mRNAs in bud",
                     limits = c(0,1))
```

