---
title: "Bud segmentation of mRNAs"
author: "Edward Wallace"
date: "2026-07-13"
output: 
  html_document:
      toc: true
---

# Summary

Analysis of CLB2 + ASH1, SUN4, SRL1, mRNAs by smFISH including bud segmentation.
Population-level cell analysis in R started by EW on 8 July 2026, continued through 13 July 2026.

Goal is to produce figures summarising the dataset:

- abundance and bud localization of mRNAs in wild-type
- comparative bud localization in different strains
- whatever cell statistics we need to support these arguments.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(ggbeeswarm)
library(cowplot)
theme_set(theme_cowplot(font_size = 10) + 
            theme(strip.background = element_blank())
)
```

# Load bud segmentation data

```{r strain_names}
# make short table here of strain ids and names in desired order
strain_table <-
  tribble(~strain, ~condition, ~strain_nice,
    "YET915", "WT",    "WT",
    "YET916", "ΔSSD1", "ssd1Δ",
    "YET919", "ΔMPT5", "mpt5∆",
    "YET563", "ΔSHE2", "she2∆",
    "YET572", "ΔSHE3", "she3∆",
    "YET918", "ΔSUN4", "sun4∆",
    "YET921", "ΔSEC72","sec72∆"
    ) %>%
  mutate(strain_nice = as_factor(strain_nice))

strain_table
```



```{r mrna_names}
mrna_names <- c("CLB2 + ASH1", "SRL1", "SUN4")

mrna_names

scale_colour_mrna <- function(...) {
  scale_colour_manual(
    values = c("CLB2 + ASH1" = "magenta3", 
               "SRL1" = "goldenrod3", 
               "SUN4" = "forestgreen"),
    ...)
}
```

```{r load_data}
mrna_alldata <- here::here("data", "Mara_budSegmentationData20260311.csv") %>%
  read_csv() %>%
  select(!starts_with("Unnamed") ) %>%
  select(!"condition + strain") %>%
  rename(row_id = "...1") %>%
  mutate(replicate = 
           factor(replicate,
                  levels = c("BR1", "BR2", "BR3", "BR4", "BR5")),
         mRNA = factor(mRNA,
                       levels = c("ASH1CLB2", "SRL1", "SUN4"), 
                       labels = c("CLB2 + ASH1", "SRL1", "SUN4"), )
         ) %>%
  rename_with(str_replace_all, pattern = "-", replacement = "_") %>%
  left_join(strain_table, by = c("strain", "condition"))

mrna_alldata
```


```{r count_by_replicate}
mrna_alldata %>% 
  filter(mRNA == "SRL1") %>%
  group_by(strain_nice, replicate) %>% 
  count() %>%
  print(n = 30)
```

# Overview plots across strains

## mRNAs per cell

```{r plot_mrna_per_cell, fig.height = 4, fig.width = 4.5}

ggplot(data = mrna_alldata) +
  geom_boxplot(aes(x = strain_nice, colour = mRNA, y = total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```


## mRNAs per mother

```{r plot_mrna_per_mother_allstrains, fig.height = 4, fig.width = 4.5}

ggplot(data = mrna_alldata, aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_mother),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```

## mRNAs per bud, filtering only for budded cells

```{r plot_mrna_per_bud_allstrains, fig.height = 4, fig.width = 4.5}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0), 
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```


## Count vs total cell area for WT 

```{r plot_countvsarea_WT, fig.height = 4, fig.width = 3.5}
plot_count_vs_area_WT <- 
  ggplot(data = mrna_alldata %>%
           filter(area > 1000,
                  strain_nice %in% c("WT")),
         aes(x = area, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "loess", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  coord_cartesian(expand = FALSE) + 
  scale_colour_mrna() +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  labs(x = "total cell area (pixels)",
       y = "mRNA count per cell") + 
  guides(colour = guide_legend(order = 1),
         linetype = "none")

plot_count_vs_area_WT
```

```{r plot_countvsarea_vsstrain, fig.width = 7}
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆"),
                area_bud > 0),
       aes(x = area, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(colour = mRNA),
              method = "gam",  se = FALSE) + 
  facet_grid(mRNA ~ strain_nice, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  coord_cartesian(xlim = c(1500,6500), expand = FALSE) +
  scale_colour_mrna()

```

## Count vs bud area for WT

```{r plot_countvsbudarea_WT, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0,
                strain_nice %in% c("WT")),
       aes(x = area_bud, y = total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "loess", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  scale_colour_mrna()  +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  labs(x = "bud area (pixels)",
       y = "mRNA count per cell")

```


## Proportion in bud, filtering only for budded cells 

```{r plot_propinbud_allstrains, fig.height = 4, fig.width = 5}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs),
               outlier.size = 0.2) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous(limits = c(0,1)) +
  scale_colour_mrna()

```

## Proportion in bud, across biological replicates, for most relevant strains 

```{r plot_propinbud_replicates, fig.height = 4, fig.width = 4}
plot_propinbud_replicates <- 
  ggplot(data = mrna_alldata %>%
           filter(area_bud > 0,
                  strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆")),
         aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs, 
                   group = interaction(strain_nice, replicate)),
               outlier.size = 0.2,
               position = position_dodge(preserve = "single")) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNAs in bud",
                     limits = c(0,1)) +
  scale_colour_mrna() + 
  labs(x = "Strain")

plot_propinbud_replicates

```


## Prop in bud vs bud size for WT

```{r plot_propinbud_WT, fig.height = 4, fig.width = 3.5}
plot_prop_vs_budarea_WT <-
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT"),
                area_bud > 0),
       aes(x = area_bud, y = total_RNAs_bud / total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(linetype = replicate, colour = mRNA),
              method = "loess", se = FALSE) + 
  facet_wrap(~ mRNA, ncol = 1, scales = "free_y") +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNA in bud", 
                     limits = c(0,1)) +
  labs(x = "Bud area (pixels)") + 
  coord_cartesian(xlim = c(0,2900), expand = FALSE) +
  scale_colour_mrna() + 
  guides(colour = guide_legend(order = 1),
         linetype = "none")

plot_prop_vs_budarea_WT
```



## Prop in bud vs bud size for relevant strains

```{r plot_propinbud_vsstrain, fig.height = 4, fig.width = 6}
ggplot(data = mrna_alldata %>%
         filter(strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆"),
                area_bud > 0),
       aes(x = area_bud, y = total_RNAs_bud / total_RNAs)) +
  geom_point(size = 0.1, colour = "grey50") +
  geom_smooth(aes(colour = mRNA),
              method = "gam",  se = FALSE) + 
  facet_grid(mRNA ~ strain_nice) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous("Proportion of mRNA in bud",
                     limits = c(0,1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1),
                     labels = c("0", "", "0.5", "", "1")) +
  labs(x = "Bud area (pixels)") + 
  coord_cartesian(xlim = c(0,2900), expand = FALSE) +
  scale_colour_mrna()

```


## Distribution of cell area by strain

```{r plot_cellareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata,
       aes(x = area)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = c(1500,6500), expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of bud area by strain

```{r plot_budareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = area_bud)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = c(0,3000), expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of mother nuclear area by strain

```{r plot_mothernucareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata ,
       aes(x = mother_nuclei_area)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  coord_cartesian(xlim = c(0,750), expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```

## Distribution of bud nuclear area by strain

Looks terrible because most buds do not have nuclei.

```{r plot_budnucareadistn, fig.height = 5, fig.width = 4}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = nuclei_bud_area)) +
  geom_density(linewidth = 1,
              kernel = "rectangular") + 
  geom_density(aes(linetype = replicate),
               colour = "grey50",
              kernel = "rectangular") + 
  facet_wrap(~ strain_nice,
             ncol = 1,
        strip.position = "left") + 
  geom_hline(yintercept = 0) + 
  #coord_cartesian(xlim = c(0,750), expand = FALSE) +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y  = element_blank(),
        axis.title.y = element_blank(),
        strip.text.y.left = element_text(angle = 0))
```
