---
title: "Bud segmentation of mRNAs"
author: "Edward Wallace"
date: "2026-07-08"
output: 
  html_document:
      toc: true
---

# Summary

Analysis of ASH1/CLB1, SUN4, SRL1, mRNAs by smFISH including bud segmentation.

This is a first attempt at analysis in R started by EW on 8 July 2026.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(ggbeeswarm)
library(cowplot)
theme_set(theme_cowplot(font_size = 10) + 
            theme(strip.background = element_blank())
)
```

# Load bud segmentation data

```{r strain_names}
# make short table here of strain ids and names in desired order
strain_table <-
  tribble(~strain, ~condition, ~strain_nice,
    "YET915", "WT",    "WT",
    "YET916", "ΔSSD1", "ssd1Δ",
    "YET919", "ΔMPT5", "mpt5∆",
    "YET563", "ΔSHE2", "she2∆",
    "YET572", "ΔSHE3", "she3∆",
    "YET918", "ΔSUN4", "sun4∆",
    "YET921", "ΔSEC72","sec72∆"
    ) %>%
  mutate(strain_nice = as_factor(strain_nice))

strain_table
```

```{r mrna_names}
mrna_names <- c("ASH1CLB2", "SRL1", "SUN4")

mrna_names

scale_colour_mrna <- function(...) {
  scale_colour_manual(
    values = c("ASH1CLB2" = "magenta3", 
               "SRL1" = "goldenrod3", 
               "SUN4" = "forestgreen"),
    ...)
}
```

```{r load_data}
mrna_alldata <- here::here("data", "Mara_budSegmentationData20260311.csv") %>%
  read_csv() %>%
  select(!starts_with("Unnamed") ) %>%
  select(!"condition + strain") %>%
  rename(cell_id = "...1") %>%
  rename_with(str_replace_all, pattern = "-", replacement = "_") %>%
  left_join(strain_table, by = c("strain", "condition"))

mrna_alldata
```

# Overview plots across strains

## mRNAs per cell

```{r plot_mrna_per_cell}

ggplot(data = mrna_alldata) +
  geom_boxplot(aes(x = strain_nice, colour = mRNA, y = total_RNAs),
               outlier.size = 0.5) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```


## mRNAs per mother

```{r plot_mrna_per_mother_allstrains}

ggplot(data = mrna_alldata, aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_mother),
               outlier.size = 0.5) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```

## mRNAs per bud, filtering only for budded cells

```{r plot_mrna_per_bud_allstrains}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0), 
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud),
               outlier.size = 0.5) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_colour_mrna()

```

## Proportion in bud, filtering only for budded cells 

```{r plot_propinbud_allstrains}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0),
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs),
               outlier.size = 0.5) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous(limits = c(0,1)) +
  scale_colour_mrna()

```

## Proportion in bud, across biological replicates, for most relevant strains 

```{r plot_propinbud_replicates}
ggplot(data = mrna_alldata %>%
         filter(area_bud > 0,
                strain_nice %in% c("WT", "ssd1Δ", "mpt5∆", "she2∆", "she3∆")),
       aes(x = strain_nice, colour = mRNA)) +
  geom_boxplot(aes(y = total_RNAs_bud / total_RNAs, 
                   group = interaction(strain_nice, replicate)),
               outlier.size = 0.5) +
  facet_wrap(~ mRNA,
             ncol = 1) +
  theme(panel.grid.major.y = element_line(colour = "grey80")) +
  scale_y_continuous(limits = c(0,1)) +
  scale_colour_mrna()

```

